A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-09-13 paper/journal.tex
Creating a first draft of the jounral version

\documentclass[runningheads]{llncs}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,pgfplots}
\usepackage[font=small]{subfig,caption}
\usepackage{algorithm2e}
\usepackage{todonotes,lineno,paralist,soul}
\usepackage{rotating,hyperref,multicol}
\usepackage{booktabs,tabularx}
\graphicspath{{figures/}}

% ==================================================================
% Format tiks and pgfplots
\tikzstyle{densely dashed}=[dash pattern=on 0.75pt off 0.5pt]
\pgfplotsset{ytick style={draw=none}}
\renewcommand{\axisdefaultheight}{185pt}
\pgfkeys{/pgf/number format/.cd,1000 sep={}}
\pgfplotscreateplotcyclelist{mylist}{%
{black,thick,smooth},
{black!50!white,thick,smooth},
{black,thick,densely dashed,smooth},
{black!50!white,thick,densely dashed,smooth},
{black,double,smooth}
}
\pgfplotsset{
        table/search path={data},
    }
\pgfplotsset{
  /pgfplots/xlabel near ticks/.style={
     /pgfplots/every axis x label/.style={
        at={(ticklabel cs:0.5)},anchor=near ticklabel
     }
  },
  /pgfplots/ylabel near ticks/.style={
     /pgfplots/every axis y label/.style={
        at={(ticklabel cs:0.5)},rotate=90,anchor=near ticklabel}
     }
}
% ==================================================================
\newcommand{\myparagraph}[1]{\smallskip\noindent\textbf{#1}.}

\author{Michael~A.~Bekos, Henry~F\"orster, Christian~Geckeler, Lukas Holl\"ander, Michael~Kaufmann, Amad\"aus~M.~Spallek, Jan~Splett}

\authorrunning{M.~A.~Bekos et al.}
\title{A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution}
\titlerunning{Drawings of Graphs with High Crossing Resolution}

\institute{
Institut f\"ur Informatik, Universit\"at T\"ubingen, T\"ubingen, Germany\\
\texttt{\{bekos,foersth,geckeler,mk\}@informatik.uni-tuebingen.de}\\
\texttt{\{jan-lukas.hollaender,amadaeus.spallek,jan.splett\} @student.uni-tuebingen.de}
}

% ==================================================================
\begin{document}
\maketitle
\linenumbers
% ==================================================================

\begin{abstract}
The \emph{crossing resolution} of a non-planar drawing of a graph is the value of the minimum angle formed by any pair of crossing edges. Recent experiments have shown that the larger the crossing resolution is, the easier it is to read and interpret a drawing of a graph. However, maximizing the crossing resolution turns out to be an NP-hard problem in general and only heuristic algorithms are known that are mainly based on appropriately adjusting force-directed algorithms.

@@

@@

In Graph Drawing, there exists a rich literature and a wide range of~techniques for drawing planar graphs; see, e.g.,~\cite{DBLP:journals/combinatorica/FraysseixPP90,DBLP:conf/gd/GutwengerM98,DBLP:journals/algorithmica/Kant96}. However, drawing a non-planar graph, and in particular when it does not have some special structure (e.g., degree restriction), is a difficult and challenging task, mainly due to the~edge crossings that negatively affect the drawing's quality~\cite{DBLP:journals/iwc/Purchase00}. As a result, the established techniques are significantly fewer (e.g., crossing minimization heuristics \cite{DBLP:journals/algorithmica/EadesW94,DBLP:journals/tsmc/SugiyamaTT81}, energy-based layout algorithms~\cite{DBLP:journals/congnum/Eades84,DBLP:journals/spe/FruchtermanR91}); for an overview refer to~\cite{DBLP:books/ph/BattistaETT99,DBLP:conf/dagstuhl/1999dg,DBLP:reference/crc/2013gd}.

In this context, Huang et al.~\cite{DBLP:conf/apvis/Huang07,DBLP:journals/vlc/HuangEH14} a decade ago introduced some important experimental evidence,
% (through eye-tracking experiments),
that edge crossings may not negatively affect the drawing's quality too much (and hence the human's ability to read and interpret it), when the angles formed by the crossing edges are large. In other words, while prior to these experiments it was commonly accepted that mainly the number of crossings is the most important parameter for judging the quality of a non-planar graph drawing, it turned out that the types of edge crossings also matter. As a result, a new and prominent research direction was initiated, recognized under the term ``beyond planarity''~\cite{Shonan2016,Dagstuhl2016,SoCG2017}, which focuses on graphs and their properties, when different constraints on the types of edges crossings are imposed; refer to~\cite{DBLP:journals/corr/abs-1804-07257} for a recent survey.

The value of the minimum angle formed by any two crossing edges in a drawing is referred to as its \emph{crossing resolution}; the crossing resolution of a graph is defined as the maximum crossing resolution over all its drawings. Clearly, the crossing resolution of a non-planar graph is at most $90^\circ$, while a graph that admits a drawing with crossing resolution $90^\circ$ is called \emph{right-angle-crossing} (\emph{RAC}) graph; see Figure~\ref{fig:examples}. Notably, RAC graphs are sparse with at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}, while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}.

\begin{figure}[t!]
	\centering
	\subfloat[\label{fig:k5} {}]{
	\includegraphics[page=1,scale=0.85]{figures/examples}}
	\hfil
	\subfloat[\label{fig:k6} {}]{
	\includegraphics[page=2,scale=0.85]{figures/examples}}
	\caption{%
	(a)~A RAC drawing of the complete graph $K_5$, and
	(b)~a drawing of the complete graph $K_6$, whose crossing resolution is arbitrarily close to $90^\circ$.}
	\label{fig:examples}
\end{figure}

The latter result is an indication that the problem of finding drawings with high crossing resolution might also be difficult, even though, formally, its complexity has not been settled yet for values of the crossing resolution smaller than $90^\circ$. Also, the literature is significantly more limited, when restricting the crossing resolution to be smaller than $90^\circ$, as also evidenced by Section~\ref{sec:relatedwork}.

From a practical point of view, we are only aware of two methods that aim at drawings with high crossing resolution; both of them are adjustments of force-directed algorithms~\cite{DBLP:journals/congnum/Eades84}. The first one is due to Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}, while the~second one is due to Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}. Common in both algorithms is that they apply appropriate forces on the endvertices of every pair of crossing edges. Each of them uses a different way to compute (the direction and the magnitude of) the forces, but the underlying idea of both is the same: the smaller the crossing angles are, the larger are the magnitudes of the forces applied at their endvertices.

In this work, we approach the crossing resolution maximization problem from a different perspective. We suggest a simple and intuitive randomization method,
% for computing drawings with high crossing resolution,
which, in a sense, mimics the way a human would try to increase the crossing resolution of a drawing. How would one increase the crossing resolution of a given drawing? First, she would try to identify the pair of edges that define the crossing resolution of the drawing (we call them \emph{critical} edges); then, she would try to move an endvertex of this pair (which we choose at random), hoping that by this move the crossing resolution will increase. Of course, we cannot consider all possible positions for the vertex to be moved. Instead, we consider a small set of randomly generated ones. If there exists a position among them, that does not lead to a reduction of the crossing resolution, we move the vertex to this~position.

In general, randomization is a technique that has not been deeply examined in Graph Drawing, as it seems difficult to even speculate about the expected quality of the produced drawings; a notable exception is the randomized approach by Goldschmidt and Takvorian~\cite{DBLP:journals/networks/GoldschmidtT94} for computing large planar subgraphs. Since we also could not provide any theoretical guarantee on the expected quality of the produced drawings,
%(mainly due to the nature of the problem itself),
we followed a more practical approach. We implemented our algorithm and the force-directed ones of~\cite{DBLP:journals/cj/ArgyriouBS13} and~\cite{DBLP:journals/vlc/HuangEHL13}, and we experimentally compared them on standard benchmark graphs.
% (such as the Rome and the North graphs) that are widely used in Graph Drawing for comparing different drawing algorithms.
Our evaluation indicates that our method significantly outperforms the force-directed ones~\cite{DBLP:journals/cj/ArgyriouBS13,DBLP:journals/vlc/HuangEHL13} in terms of crossing resolution, but this comes at the cost of slightly worse running time
%, especially
for large and dense graphs. Analogous results are obtained, when our algorithm and the ones of~\cite{DBLP:journals/cj/ArgyriouBS13} and~\cite{DBLP:journals/vlc/HuangEHL13} are adjusted to maximize  the \emph{angular resolution} (i.e., the minimum value of the angle between any two adjacent edges~\cite{DBLP:journals/siamcomp/FormannHHKLSWW93}) or the \emph{total resolution} (i.e., the minimum of the angular and the crossing resolution~\cite{DBLP:journals/cj/ArgyriouBS13}).

\medskip\noindent{\it Preliminaries:}
Unless otherwise specified, in this paper we consider simple undirected graphs. Let $G=(V,E)$ be such a graph. The degree of vertex $u\in V$ of $G$ is denoted by $d(u)$. The degree $d(G)$ of  graph $G$ is defined as the maximum degree of its vertices, i.e., $d(G)=\max_{u\in V}d(u)$.
%
Given a drawing $\Gamma(G)$ of $G$, we denote by $p(u)=(x_u,y_u)$ the position of vertex $u \in V$ of $G$ in $\Gamma(G)$. %For a pair of vertices $u,v\in V$ of $G$, the unit-length vector from $p(u)$ to $p(v)$ is denoted as $\overrightarrow{p(u)p(v)}$, while the line segment delimited by $p(u)$ and $p(v)$ as $\overline{p(u)p(v)}$.

\medskip\noindent{\it Structure of the paper:}
The remainder of this paper is structured as follows. Section~\ref{sec:relatedwork} overviews related works.
%In Section~\ref{sec:preliminaries}, we introduce preliminary notions and the notation used throughout the paper.
Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} in Section~\ref{sec:experiments}, where we also discuss our insights from this project. In the appendix, we provide experimental results on grid restricted drawings, on more test sets and on the graphs from the Graph Drawing Competition in 2017.

@@

In this section, we present the results of our experimental evaluation. For comparison purposes, apart from our algorithm, we also implemented the force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. The implementations\footnote{Our implementation is available on request from the authors.} were in Java using yFiles~\cite{DBLP:books/sp/04/WieseE004}. The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM.
%
As a test set for our experiment, we used the non-planar Rome graphs~\cite{DBLP:reference/crc/BattistaD13}, which form a collection of around 8.100 benchmark graphs; in Appendix~\ref{app:north}, we also report on the AT\&T graphs.

The experiment was performed as follows. Initially, each Rome graph was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with
\begin{inparaenum}[(i)]
\item our algorithm,
\item our algorithm restricted~not to violate the aspect ratio of the initial layout, and the force-directed algorithms
\item by Argyriou et al.\ and
\item by Huang et al.
\end{inparaenum}
%We compared the quality of the produced drawings based on the following aesthetic~properties:
%
%\begin{multicols}{2}
%\begin{enumerate}[P.1.]
%\item \label{p:cr-res} crossing resolution
%\item \label{p:to-res} total resolution
%\item \label{p:an-res} angular resolution
%\columnbreak
%\item \label{p:as-rat} aspect ratio
%\item \label{p:no-xing} number of crossings
%\end{enumerate}
%\end{multicols}
%
Since all algorithms of the experiment can easily be adjusted to maximize~only the crossing resolution, or only the angular resolution or both (by maximizing~the total resolution), we adjusted each of the algorithms to maximize exclusively the corresponding measures; see Figs.~\ref{fig:cr-res}, \ref{fig:to-res} and~\ref{fig:an-res}. In our algorithm, this can be achieved by modifying appropriately the content of the vertex-pool (as we saw in Section~\ref{ssec:variants}), while in the algorithms of Argyriou et al.\ and of Huang et al.\ by switching on only the forces that maximize the corresponding properties under measure (note that, each of these two algorithms has a different set of forces to maximize the crossing and the angular resolution, such that together they maximize the total resolution). The reported results are on average across different drawings with same number of vertices. Finally, we mention that for our algorithm, we chose $\delta_{max}=\frac{1}{2}\max\{w,h\}$, where $w$ and $h$ are the width and the height of the initial drawing, respectively, $\delta_{min}=\frac{1}{100}\delta_{max}$ and $\rho=10$.

%We compared the quality of the produced drawings based on crossing resolution, total resolution, and angular resolution. Additionally, we consider the aspect ratio, and the number of crossings of the corresponding drawings.

@@

\begin{figure}[t]
\centering
\subfloat[\label{fig:cr-res-1}{Crossing resolution vs no.~of vertices}]{
\centering
\scalebox{0.95}{
\centering
%======================
% Crossing Resolution
%======================
\begin{tikzpicture}
\begin{axis}[axis x line*=bottom,xlabel style={yshift=0.2cm},axis y line*=left,ylabel style={yshift=-0.2cm}, legend style={at={(0.425,1.35)},anchor=north,legend columns=2,draw=none}, width=0.475\textwidth, cycle list name=mylist,
mark repeat={5}, xtick={10,20,30,40,50,60,70,80,90,100}, ytick={0,10,20,30,40,50,60,70,80,90}, ylabel={Crossing Resolution}, xlabel={Number of Vertices}, xmin=10, xmax=100, tick pos=left,ymajorgrids]
\addplot table [x=n, y=crossing resolution rm, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=n, y=crossing resolution rm-1, col sep=semicolon] {crossingResolution.csv};
\addplot table [x=n, y=crossing resolution fa, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=Nodes, y=Crossing Only, col sep=semicolon]{abs-crossingResolution.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}}
\subfloat[\label{fig:cr-res-2}{Aspect ratio vs no.~of vertices}]{
\centering
\scalebox{0.95}{
\centering
%===================================
% Crossing Resolution Aspect Ratio
%===================================
\begin{tikzpicture}
\begin{axis}[axis x line*=bottom,xlabel style={yshift=0.2cm},axis y line*=left,ylabel style={yshift=-0.2cm},legend style={at={(0.425,1.35)}, anchor=north,legend columns=2,draw=none}, width=0.475\textwidth,cycle list name=mylist, mark repeat={5},xtick={10,20,30,40,50,60,70,80,90,100},ylabel={Aspect Ratio}, xlabel={Number of Vertices}, ytick={1,2.5,10,25,100,250,1000,2500,10000}, xmin=10, xmax=100,ymode=log, log ticks with fixed point,tick pos=left, ymajorgrids,]
\addplot table [x=n, y=Aspect ratio rm, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=n, y=Aspect ratio rm-1, col sep=semicolon] {crossingResolution.csv};
\addplot table [x=n, y=Aspect ratio fa, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=Nodes, y=Crossing Only, col sep=semicolon]{abs-aspectRatio.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}}

>>>>

\documentclass{comjnl}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,pgfplots}
\usepackage[font=small]{subfig,caption}
%\usepackage{algorithm2e}
\usepackage{todonotes,lineno,paralist,soul}
\usepackage{rotating,hyperref,multicol}
\usepackage{booktabs,tabularx}
\graphicspath{{figures/}}

% ==================================================================
% Format tiks and pgfplots
\tikzstyle{densely dashed}=[dash pattern=on 0.75pt off 0.5pt]
\pgfplotsset{ytick style={draw=none}}
\renewcommand{\axisdefaultheight}{185pt}
\pgfkeys{/pgf/number format/.cd,1000 sep={}}
\pgfplotscreateplotcyclelist{mylist}{%
{black,thick,smooth},
{black!50!white,thick,smooth},
{black,thick,densely dashed,smooth},
{black!50!white,thick,densely dashed,smooth},
{black,double,smooth}
}
\pgfplotsset{
        table/search path={data},
    }
\pgfplotsset{
  /pgfplots/xlabel near ticks/.style={
     /pgfplots/every axis x label/.style={
        at={(ticklabel cs:0.5)},anchor=near ticklabel
     }
  },
  /pgfplots/ylabel near ticks/.style={
     /pgfplots/every axis y label/.style={
        at={(ticklabel cs:0.5)},rotate=90,anchor=near ticklabel}
     }
}
% ==================================================================
\newcommand{\myparagraph}[1]{\smallskip\noindent\textbf{#1}.}
% ==================================================================

\begin{document}

\author{Michael~A.~Bekos} 
\author{Henry~F\"orster}
\author{Christian~Geckeler} 
\author{Lukas~Holl\"ander} 
\author{Michael~Kaufmann}
\author{Amad\"aus~M.~Spallek} 
\author{Jan~Splett}

\title[Drawings of Graphs with High Crossing Resolution]{A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution}

\affiliation{
Institut f\"ur Informatik, Universit\"at T\"ubingen, T\"ubingen, Germany\\
}

\email{\{bekos,foersth,mk\}@informatik.uni-tuebingen.de\\\{geckeler,jan-lukas.hollaender,amadaeus.spallek,jan.splett\}@student.uni-tuebingen.de}

\received{13 September 2018}

\shortauthors{M.~A.~Bekos et al.}

\keywords{Graph Drawing, Angular, Crossing and Total Resolution.}

%\copyrightyear{2018} \vol{00} \issue{0} \DOI{000}
% ==================================================================

\begin{abstract}
The \emph{crossing resolution} of a non-planar drawing of a graph is the value of the minimum angle formed by any pair of crossing edges. Recent experiments suggest that the larger the crossing resolution is, the easier it is to read and interpret a drawing of a graph. However, maximizing the crossing resolution turns out to be an NP-hard problem in general, and only heuristic algorithms are known that are mainly based on appropriately adjusting force-directed algorithms.

@@

% ==================================================================
\maketitle
% ==================================================================

@@

In Graph Drawing, there exists a rich literature and a wide range of techniques for drawing planar graphs; see, e.g.,~\cite{DBLP:journals/combinatorica/FraysseixPP90,DBLP:conf/gd/GutwengerM98,DBLP:journals/algorithmica/Kant96}. However, drawing a non-planar graph, and in particular when it does not have some special structure (e.g., degree restriction), is a quite difficult and very challenging task, mainly due to the edge crossings that negatively affect the drawing's quality~\cite{DBLP:journals/iwc/Purchase00}. As a result, the established techniques are significantly fewer (e.g., crossing minimization heuristics~\cite{DBLP:journals/algorithmica/EadesW94,DBLP:journals/tsmc/SugiyamaTT81}, energy-based layout algorithms~\cite{DBLP:journals/congnum/Eades84,DBLP:journals/spe/FruchtermanR91}); for an overview we point the reader to~\cite{DBLP:books/ph/BattistaETT99,DBLP:conf/dagstuhl/1999dg,DBLP:reference/crc/2013gd}.

In this context, Huang et al.~\cite{DBLP:conf/apvis/Huang07,DBLP:journals/vlc/HuangEH14} a decade ago introduced some important experimental evidence (through eye-tracking experiments), that edge crossings may not negatively affect the drawing's quality too much (and hence the human's ability to read and interpret it), when the angles formed by the crossing edges are large. In other words, while prior to these experiments it was commonly accepted that mainly the number of crossings is the most important parameter for judging the quality of a non-planar graph drawing~\cite{DBLP:conf/gd/Purchase97,DBLP:journals/iwc/Purchase00}, it turned out that the types of edge crossings also matter. As a result, a new and prominent research direction was initiated, recognized under the term ``beyond planarity''~\cite{Shonan2016,Dagstuhl2016,SoCG2017}, which focuses on graphs and their properties, when different constraints on the types of edges crossings are imposed; refer to~\cite{DBLP:journals/corr/abs-1804-07257} for a recent survey.

Formally, the value of the minimum angle formed by any two crossing edges in a drawing is referred to as its \emph{crossing resolution\footnote{Also, refered to as \emph{crossing angle resolution} in the literature.}}. Analogously, the crossing resolution of a graph is defined as the maximum crossing resolution over all its drawings. Clearly, the crossing resolution of a non-planar graph is at most $90^\circ$, while a graph that admits a drawing with crossing resolution $90^\circ$ is called \emph{right-angle-crossing} (or \emph{RAC}, for short) graph; see Figure~\ref{fig:examples} for an illustration. Notably, RAC graphs are sparse (an $n$-vertex RAC graph has at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}), while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}.

\begin{figure}[t!]
	\centering
	\subfloat[\label{fig:k5} {}]{
	\includegraphics[page=1]{figures/examples}}
	\hfil
	\subfloat[\label{fig:k6} {}]{
	\includegraphics[page=2]{figures/examples}}
	\caption{Illustration of: %
	(a)~a RAC drawing of the complete graph $K_5$, and
	(b)~a drawing of the complete graph $K_6$, whose crossing resolution is arbitrarily close to $90^\circ$.}
	\label{fig:examples}
\end{figure}

The latter result is an indication that the problem of finding drawings with high crossing resolution might also be difficult, even though, formally, its complexity has not been settled yet for values of the crossing resolution smaller than $90^\circ$. It is also worth noting that the literature is significantly more limited, when restricting the crossing resolution to be smaller than $90^\circ$, as also evidenced by Section~\ref{sec:relatedwork}.

From a practical point of view, we are only aware of two methods that aim at drawings with high crossing resolution; both of them are adjustments of force-directed algorithms~\cite{DBLP:journals/congnum/Eades84}. The first one is due to Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}, while the~second one is due to Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. Common in both algorithms is that they apply appropriate forces on the endvertices of every pair of crossing edges. Each of them uses a different way to compute (the direction and the magnitude of) the forces, but the underlying idea of both is the same: the smaller the crossing angles are, the larger are the magnitudes of the forces applied at their endvertices.

In this work, we approach the crossing resolution maximization problem from a different perspective. We suggest a simple and intuitive randomization method for computing drawings with high crossing resolution, which, in a sense, mimics the way a human would try to increase the crossing resolution of a drawing. How would one increase the crossing resolution of a given drawing? First, she would try to identify the pair of edges that define the crossing resolution of the drawing (we call them \emph{critical} edges); then, she would try to move an endvertex of this pair (which we choose at random), hoping that by this move the crossing resolution will increase. Of course, we cannot consider all possible positions for the vertex to be moved. Instead, we consider a small set of randomly generated ones. If there exists a position among them, that does not lead to a reduction of the crossing resolution, we move the vertex to this position.

In general, randomization is a technique that has not been deeply examined in Graph Drawing, as it seems difficult to even speculate about the expected quality of the produced drawings; a notable exception is the randomized approach by Goldschmidt and Takvorian~\cite{DBLP:journals/networks/GoldschmidtT94} for computing large planar subgraphs. Since we also could not provide any theoretical guarantee on the expected quality of the produced drawings (mainly due to the nature of the problem itself),
we followed a more practical approach. We implemented our algorithm and the force-directed ones of~\cite{DBLP:journals/cj/ArgyriouBS13} and~\cite{DBLP:journals/vlc/HuangEHL13}, and we experimentally compared them on standard benchmark graphs that are widely used in Graph Drawing for comparing different drawing algorithms. Our evaluation indicates that our method significantly outperforms the force-directed algorithms~\cite{DBLP:journals/cj/ArgyriouBS13,DBLP:journals/vlc/HuangEHL13} in terms of crossing resolution, but this comes at the cost of slightly worse running time, especially for large and dense graphs. Analogous results are obtained, when our algorithm and the ones of~\cite{DBLP:journals/cj/ArgyriouBS13} and~\cite{DBLP:journals/vlc/HuangEHL13} are adjusted to maximize the \emph{angular resolution} (that is, the minimum value of the angle between any two adjacent edges~\cite{DBLP:journals/siamcomp/FormannHHKLSWW93}) or the \emph{total resolution} (that is, the minimum of the angular and the crossing resolution~\cite{DBLP:journals/cj/ArgyriouBS13}).

\medskip\noindent{\it Preliminaries:}
Unless otherwise specified, in this paper we consider simple undirected graphs. Let $G=(V,E)$ be such a graph. The degree of vertex $u\in V$ of $G$ is denoted by $d(u)$. The degree $d(G)$ of  graph $G$ is defined as the maximum degree of its vertices, i.e., $d(G)=\max_{u\in V}d(u)$. Given a drawing $\Gamma(G)$ of $G$, we denote by $p(u)=(x_u,y_u)$ the position of vertex $u \in V$ of $G$ in $\Gamma(G)$. %For a pair of vertices $u,v\in V$ of $G$, the unit-length vector from $p(u)$ to $p(v)$ is denoted as $\overrightarrow{p(u)p(v)}$, while the line segment delimited by $p(u)$ and $p(v)$ as $\overline{p(u)p(v)}$.

\medskip\noindent{\it Structure of the paper:}
The remainder of this paper is structured as follows. Section~\ref{sec:relatedwork} overviews related works. Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} in Section~\ref{sec:experiments}, where we also discuss our insights from this project.

@@

In this section, we present the results of our experimental evaluation. For comparison purposes, apart from our algorithm, we also implemented the force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. The implementations\footnote{Our implementation is available on request from the authors.} were in Java using yFiles~\cite{DBLP:books/sp/04/WieseE004}. The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM. As test sets for our experiment, we used the non-planar Rome graphs~\cite{DBLP:reference/crc/BattistaD13} and the AT\&T graphs, which are two collections of around 8.100 and 424 benchmark graphs, respectively.

The experiment was performed as follows. Initially, each graph of our expertiment was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with
\begin{inparaenum}[(i)]
\item our algorithm,
\item our algorithm restricted not to violate the aspect ratio of the initial layout, and the force-directed algorithms
\item by Argyriou et al.\ and
\item by Huang et al.
\end{inparaenum}
%We compared the quality of the produced drawings based on the following aesthetic properties:
%
%\begin{enumerate}[P.1.]
%\item \label{p:cr-res} crossing resolution
%\item \label{p:to-res} total resolution
%\item \label{p:an-res} angular resolution
%\item \label{p:as-rat} aspect ratio
%\item \label{p:no-xing} number of crossings
%\end{enumerate}

Since all algorithms of the experiment can easily be adjusted to maximize~only the crossing resolution, or only the angular resolution or both (by maximizing~the total resolution), we adjusted each of the algorithms to maximize exclusively the corresponding measures; see Figs.~\ref{fig:cr-res}, \ref{fig:to-res} and~\ref{fig:an-res}. In our algorithm, this can be achieved by modifying appropriately the content of the vertex-pool (as we saw in Section~\ref{ssec:variants}), while in the algorithms of Argyriou et al.\ and of Huang et al.\ by switching on only the forces that maximize the corresponding properties under measure (note that, each of these two algorithms has a different set of forces to maximize the crossing and the angular resolution, such that together they maximize the total resolution). The reported results are on average across different drawings with same number of vertices. Finally, we mention that for our algorithm, we chose $\delta_{max}=\frac{1}{2}\max\{w,h\}$, where $w$ and $h$ are the width and the height of the initial drawing, respectively, $\delta_{min}=\frac{1}{100}\delta_{max}$ and $\rho=10$.

\subsection{The Rome Graph Test Set}

In this subsection, we report the results of our expertimental evaluation for the non-planar Rome graphs, which form a collection of 424 benchmark graphs. Note that we did not impose any grid constraint on our algorithms. Our results are summarized in Figs.~\ref{fig:cr-res}, \ref{fig:an-res} and~\ref{fig:to-res}.

@@

\begin{figure*}[t]
\centering
\subfloat[\label{fig:cr-res-1}{Crossing resolution vs no.~of vertices}]{
\centering
\scalebox{0.95}{
\centering
%======================
% Crossing Resolution
%======================
\begin{tikzpicture}
\begin{axis}[axis x line*=bottom,xlabel style={yshift=0.2cm},axis y line*=left,ylabel style={yshift=-0.2cm}, legend style={at={(0.425,1.35)},anchor=north,legend columns=2,draw=none}, width=0.475\textwidth, cycle list name=mylist,
mark repeat={5}, xtick={10,20,30,40,50,60,70,80,90,100}, ytick={0,10,20,30,40,50,60,70,80,90}, ylabel={Crossing Resolution}, xlabel={Number of Vertices}, xmin=10, xmax=100, tick pos=left,ymajorgrids]
\addplot table [x=n, y=crossing resolution rm, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=n, y=crossing resolution rm-1, col sep=semicolon] {crossingResolution.csv};
\addplot table [x=n, y=crossing resolution fa, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=Nodes, y=Crossing Only, col sep=semicolon]{abs-crossingResolution.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}}
\subfloat[\label{fig:cr-res-2}{Aspect ratio vs no.~of vertices}]{
\centering
\scalebox{0.95}{
\centering
%===================================
% Crossing Resolution Aspect Ratio
%===================================
\begin{tikzpicture}
\begin{axis}[axis x line*=bottom,xlabel style={yshift=0.2cm},axis y line*=left,ylabel style={yshift=-0.2cm},legend style={at={(0.425,1.35)}, anchor=north,legend columns=2,draw=none}, width=0.475\textwidth,cycle list name=mylist, mark repeat={5},xtick={10,20,30,40,50,60,70,80,90,100},ylabel={Aspect Ratio}, xlabel={Number of Vertices}, ytick={1,2.5,10,25,100,250,1000,2500,10000}, xmin=10, xmax=100,ymode=log, log ticks with fixed point,tick pos=left, ymajorgrids,]
\addplot table [x=n, y=Aspect ratio rm, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=n, y=Aspect ratio rm-1, col sep=semicolon] {crossingResolution.csv};
\addplot table [x=n, y=Aspect ratio fa, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=Nodes, y=Crossing Only, col sep=semicolon]{abs-aspectRatio.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}}

\subfloat[\label{fig:cr-res-3}{No.~of crossings vs no.~of vertices}]{
\centering
\scalebox{0.95}{
\centering
%======================================
% Crossing Resolution Crossing Number
%======================================
\begin{tikzpicture}
\begin{axis}[axis x line*=bottom,xlabel style={yshift=0.2cm},axis y line*=left,ylabel style={yshift=-0.2cm},legend style={at={(0.425,1.35)}, anchor=north,legend columns=2,draw=none}, width=0.475\textwidth,cycle list name=mylist,  mark repeat={5},xtick={10,20,30,40,50,60,70,80,90,100}, xmin=10, xmax=100, ytick={0,25,50,75,100,125,150,175,200},ylabel={Number of Crossings}, xlabel={Number of Vertices},tick pos=left, ymajorgrids]
\addplot table [x=n, y=Crossing number rm, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=n, y=Crossing number rm-1, col sep=semicolon] {crossingResolution.csv};
\addplot table [x=n, y=Crossing number fa, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=Nodes, y=Crossing Only, col sep=semicolon]{abs-crossings.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}}
\subfloat[\label{fig:cr-res-4}{No.~of iterations vs no.~of vertices}]{
\centering
\scalebox{0.95}{
\centering
%==================================
% Crossing Resolution Iterations
%==================================
\begin{tikzpicture}
\begin{axis}[axis x line*=bottom,xlabel style={yshift=0.2cm},axis y line*=left,ylabel style={yshift=-0.2cm},legend style={at={(0.425,1.35)}, anchor=north,legend columns=2,draw=none}, width=0.475\textwidth,cycle list name=mylist,  ylabel={Iterations}, xlabel={Number of Vertices}, mark repeat={5},xtick={10,20,30,40,50,60,70,80,90,100}, xmin=10,
xmax=100, ytick={0,500,1000,1500,2000,2500,3000,3500,4000,4500,5000},tick pos=left, ymajorgrids]
\addplot table [x=n, y=Iterations rm, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=n, y=Iterations rm-1, col sep=semicolon] {crossingResolution.csv};
\addplot table [x=n, y=Iterations fa, col sep=semicolon]{crossingResolution.csv};
\addplot table [x=Nodes, y=Crossing Only, col sep=semicolon]{abs-iterations.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}}
\caption{Experimental results on the crossing resolution for the Rome graphs.}
\label{fig:cr-res}
\end{figure*}
